---
title: "Class 12"
author: "Josefina O'Toole (PID: A16978557)"
format: pdf
---

# Section 1. Proportion of G|G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39759942-40030277;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G

```{r}
round( table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2 )
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Lets now dig into this further.

## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. 

Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

> Q13: The sample size for each genotype is listed below.

```{r}
table(expr$geno)
```

```{r}
summary(expr)
```

```{r}
library(ggplot2)
```

Lets make a boxplot

```{r}
bplot <- ggplot(expr) +
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
bplot
```

> Q14: Lower expression correlates with the G|G genotype.


