---
title: "Class 18"
author: "Josefina O'Toole (PID: A16978557)"
format: pdf
---

Pertussis (a.k.a.) Whooping Cough is a deadly lung infection caussed by the bacteria B. Pertussis.

The CDC tracks Pertussis cases around the US.
http://tinyurl.com/pertussiscdc

We can "scrape" this data using the R **datapasta** package.

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r, include=FALSE}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2024L
  ), cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    35493
  )
       )

```

```{r}
head(cdc)
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_line() +
  geom_vline(xintercept = 1946, col="blue") +
  geom_vline(xintercept = 1996, col="red") +
  geom_vline(xintercept = 2020, col="green") +
  geom_vline(xintercept = 2003, col="purple")
```

There were high cases numbers before the first wP (whole-cell) vaccine roll out in 1946 then a rapid decline in case numbers until 2004 when we have our first large-scale outbreaks of pertussis again. There is also a notable COVID related dip and recent rapid rise.

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There is a lag, a waning of immunity if you have the aP vaccine. There is an initial decrease in cases and then a resurgence. The immunity generated by the acellular pertussis vaccine is not as long-lasting as that from the whole-cell vaccine.

What is different about the immune response to infection if you had an older wP vaccine v.s. the newer aP vaccine?

## Computational Models of Immunity Pertussis Boost (CMI-PB)

The CMI-PB project aims to address this key question: what is different between aP and wP individuals.

We can get all the data from this ongoing project via JSON API calls.
For this we will use the **jsonlite** package. We can install with: `install.packages("jsonlite")`

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subject)
```

> How many individual "subjects" are in this dataset?

```{r}
nrow(subject)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
today() - ymd("2000-01-01")
```

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```

```{r}
# Use todays date to calculate age in days
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

(i) 36
(ii) 27
(iii) They are significantly different.

```{r}
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```


> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

I think these two groups are significantly different.

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```


This is not representative of the US population but it is the biggest dataset of its type so let's see what we can learn:

Obtain more data from CMI-PB...

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = T)

ab_data <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = T)
```

```{r}
head(specimen)
```

```{r}
head(ab_data)
```

I now have 3 tables of data from CMI-PB: `subject`, `specimen`, and `ab_data`. I need to "join" these tables so I will have all the info I need to work with.

For this we will use the `inner_join()` function from the **dplyr** package.

> Q9.2

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)

head(meta)
```

```{r}
dim(subject)
dim(specimen)
```

```{r}
dim(meta)
```


Now we can join our `ab_data` table to `meta` so we have all the info we need about antibody levels.

```{r}
abdata <- inner_join(meta, ab_data)

head(abdata)
```

> How many different antibody isotypes are there in this dataset?

```{r}
length(abdata$isotype)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?


```{r}
table(abdata$isotype)
```

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

The number of rows on the most "recent" dataset is half as large as the number of rows from 2020 but twice as large as the number of rows for 2021 and 2022.

```{r}
table(abdata$dataset)
```


```{r}
table(abdata$antigen)
```

I want a plot of antigen levels across the whole dataset.

```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```

```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

Antigens like FIM2/3, PT, FELD1 have quite a large range of values. Others like Measles don't show much activity.

```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

## Examine IgG Ab titer levels

For this I need to select out just isotype IgG.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

A overview boxplot:
```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```


>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The Fim2/3 antigen, a component of acellular pertussis vaccines, refers to the fimbriae (Fim) types 2 and 3 produced by Bordetella pertussis, the bacteria causing whooping cough, which are important for bacterial adhesion and are used as vaccine antigens.


Digging in further to look at the course of IgG isotype PT antigen levels across aP and wP individuals:

```{r}
## Filter to include 2021 data only
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

## Filter to look at IgG PT data only
abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  
  #Plot and color by infancy_vac (wP vs aP)
ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```


> Q18. Does this trend look similar for the 2020 dataset?





