---
title: "Class 6 Quesetion 6"
author: "Josefina O'Toole (PID: A16978557)"
format: pdf
---

> Q6. Write a function: See Q6 of the hands-on lab supplement above. This entails turning a supplied code snippet into a more robust and re-usable function that will take any of the three listed input proteins and plot the effect of drug binding. Note assessment rubric and submission instructions within document.

# Can you improve this analysis code?
```{r}
library(bio3d)

s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")

```

There was a bug so I changed s1 to s3.

Shorted code:

```{r}
library(bio3d)

functionQ6 <- function(protein) {
  
  ad <- read.pdb(protein) # this line reads the protein pdb file
  # ad stands for any database
  
  ad.chainA <- trim.pdb(ad, chain="A", elety="CA") # this line takes the chainA
  # from the database and trims it to a smaller subset of atoms
  
  ad.b <- ad.chainA$atom$b # this line assigns the ad.chainA from above 
  # that have the b column to a new variable called ad.b
  
  plotb3(ad.b, sse=ad.chainA, typ="l", ylab="Bfactor") # this line plots the 
  # numbers in the ad.b vector
}

# the input of the function is a pdb file of a specific protein
functionQ6("4AKE")
functionQ6("1AKE")
functionQ6("1E4Y")
# the output is a standard scatter plot with a line, of the chainA & b category
# of the protein with Beta factor on the y-axis and the residues of the protein
# on the x-axis.

```



